Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2017-05-12 pkg/glmertree/paper/treatment_interactions.Rnw
Some paper revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1327 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{bigfoot}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}
\usepackage{Sweave}
\usepackage{subfig}

@@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees. Also, GLMM tree show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets. An implementation is available in the R package \verb|glmertree| and a short tutorial demonstrating how to fit GLMM trees in practice is provided in the accompanying vignette.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods. Tree-based methods, also known as recursive partitioning methods, recursively split the space spanned by the predictor variables into rectangular regions, containing observations that are increasingly similar with respect to the outcome. Several tree-based methods take the mean or majority class of a single dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}. Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.

Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the splitting variable duration and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3. As the observations in node~2 are not further split, node~2 is a terminal node. Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.

If the partition in Figure~\ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables. The prediction is then determined by the distribution of the training observations within the terminal node. For example, the prediction may be the node-specific mean of a single continuous variable. In the current paper, we focus on trees where the terminal nodes consist of a linear (LM) or generalized linear model (GLM), in which case the predicted value for a new observation is determined by the node-specific parameter estimates of the (G)LM, while also adjusting for random effects.

@@

Tree-based methods are particularly useful for exploratory purposes, because they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, they are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio} and \citeA{ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

@@

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects \cite<e.g., >{HajjyBell11,SelaySimo12}, GLMM trees allow for treatment effect estimation, with continuous as well as non-continuous response variables.

The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions. The second section is a short tutorial on how to fit GLMM trees using the R package \verb|glmertree| \readme{still needs to be adapted: vignette}. In the third section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions. In the fourth section, we apply a GLMM tree to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. In the fith and last section we summarize the results and discuss limitations and directions for future research.

@@

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. The R code used to generate the example dataset is provided in the Appendix \readme{I would put this into the appendix of the vignette instead.}.

The dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Finally, each patient was part of one of ten clusters, each having a different value for the random intercept, uncorrelated with the partitioning variables.

The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2. For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions. For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1. Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not. Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions. 
\readme{I've incorporated a characterization of the groups in terms of duration and anxiety here. But probably a better description of these variables is needed earlier. Possibly in a table with some summary statistics, e.g., mean/sd/range. Finally, the description in the manual Rd pages of the package should also be improved.}

@@

The rationale behind MOB is that a single global GLM (or other parametric model) may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition.
For example, to assess the effect of treatment, we may first fit a global GLM where the treatment indicator has the same effect/coefficient on the outcome for all observations. Subsequently, the data may be partitioned recursively with respect to other covariates, leading to separate models with different treatment effects/coefficients in each subsample.
\readme{Expanded on the treatment GLM MOB explanation to give more context and motivation.}

@@

The MOB algorithm can be used to partition the dataset using additional covariates and find better-fitting local models. To this end, the MOB algorithm tests for parameter stability over a set of auxiliary covariates, also called \textit{partitioning variables}. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

@@

If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{{k^*}}$ \readme{the U variables have not been introduced, no symbol has been used for the partitioning variables so far}, the partitioning variable with the minimal $p$-value in step (2). The split point for $U_{k^*}$ is selected by taking the value that minimizes the sum of the values of the objective function in both partitions. In other words, for every possible split point, the objective function is minimized separately in the two resulting subsets of the data; the split point yielding the minimum sum of the objective functions is selected. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a mixed-effects model would be more appropriate. The GLM in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a random-intercept only model, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element corresponding to the random intercept for cluster $m$. For simplicity, we employ a cluster-specific intercept only, but further random effects can easily be included in $z_i$. Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. Also, the errors $\epsilon$ are assumed to follow a normal distribution, with constant variance across clusters. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

Although the random-effects part of the GLMM in Equation~\ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation~\ref{eq:fixedeffects_MOB} while still adjusting for random effects: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects parameters are estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random-effects parameters.

@@

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the GLM tree is re-estimated in step (1) and the fixed- and random-effects parameters are re-estimated in step (2). Note that the random-effects part of the model is not partitioned, but estimated globally. Only the fixed effects are estimated locally, within the cells of the partition. Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}. Typically, this converges if the tree does not change from one iteration to the next.

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. In addition to the treatment indicator, treatment outcome and the potential partitioning variables, the GLMM tree has also included a random intercept term with respect to the cluster indicator. As a result, the dependence between observations is taken into accoutn, the true treatment subgroups have been recovered and the spurious split involving the anxiety variable no longer appears in the tree.

@@

\section{Tutorial}

\readme{This tutorial should be turned into a vignette and then the R output, graphics, etc. should be shown as well.}

We implemented the GLMM tree algorithm in the R package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). The package makes use of the \verb|partykit| package \cite{HothyZeil15} and \verb|lme4| package \cite{BateyMeac12} to estimate trees and mixed-effects models, respectively. The latest version of the \verb|glmertree| package can be downloaded, installed and loaded as follows:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
install.packages("glmertree", repos="http://R-Forge.R-project.org")
library(glmertree)
@
\end{small}

Package documentation and examples can be accessed as follows:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
?glmertree
@
\end{small}

The main functions in the \verb|glmertree| package are \verb|lmertree()|, for continuous outcome variables, and \verb|glmertree()|, for binary or count outcome variables. Both functions require the user to specify at least two arguments: \verb|formula| and \verb|data|. The data we will be using is the motivating example dataset described in the previous section. This dataset is included in the \verb|partykit| package and can be loaded as follows:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
data("a_m_data", package = "glmertree")
summary(a_m_data)
@
\end{small}

The model formula to be specified consists of a left- and right hand side. The left hand side of the model formula (preceding the tilde symbol) specifies the outcome variable. The right hand side consists of three parts, separated by vertical bars: The first part specifies the predictor variable(s) of the (generalized) linear model, the second part specifies the random effects and the third part specifies the potential partitioning variables:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
exampleglmmtree <- lmertree(depression ~ treatment | cluster | age + duration + 
                            anxiety, data = a_m_data)
@
\end{small}

Note that in the example above, the partitioning variables are continuous, but (ordered) categorical partitioning variables may also be specified. Also, we specified only a single variable in the random-effects part, resulting in estimation of a random intercept with respect to \verb|cluster|. More complex random effects can also be specified: for example, specifying the random-effects part as \verb=... | (1 + age | cluster) | ...= would yield a model with a random intercept as well as a random slope for age with respect to cluster. The brackets are necessary to protect the vertical bars in the random effects formulation.

Now, we can use the \verb|plot| method to recreate the tree depicted in Figure~\ref{fig:example_glimmertree}:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
plot(exampleglmmtree)
@
\end{small}

In every inner node of the tree, the splitting variable and corresponding p-value from the parameter stability test is reported. To control for multiple testing, the p-values are Bonferroni corrected, by default. This can be turned off by adding \verb|bonferroni = FALSE| to the function call, yielding a less conservative criterion for the parameter stability tests, but note that this will increase the likelihood of overfitting. The significance level $\alpha$ equals .05 by default, but a different value, say for example .01, can be specified by including \verb|alpha = .01| in the function call.

The predictions of the random effects can be plotted by adding \verb|plotranef = TRUE| to the function call:
\readme{As already indicated by e-mail, plotranef is an awkward name and should be improved, e.g., to which.}

\begin{small}
<<eval = FALSE, echo = TRUE>>=
plot(exampleglmmtree, plotranef = TRUE)
@
\end{small}

To obtain numerical results, \verb|print|, \verb|coef| and \verb|ranef| methods can be used:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
print(exampleglmmtree)
coef(exampleglmmtree)
ranef(exampleglmmtree)
@
\end{small}

To obtain predicted values, the \verb|predict| method can be used:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
predict(exampleglmmtree, newdata = a_m_data[1:10,])
@
\end{small}

When \verb|newdata| is not specified, predictions for the training observations are returned, by default. Random effects can be excluded from the predictions by adding \verb|re.form = NA|. This is useful, for example, when \verb|newdata| is specified, but the new observations do not have a cluster indicator or are from new clusters:

\begin{small}
<<eval = FALSE, echo = TRUE>>=
predict(exampleglmmtree, newdata = a_m_data[1:10, -4], re.form = NA)
@
\end{small}

Residuals of the fitted GLMM tree can be obtained with the \verb|residuals| method. This can be useful for assessing potential misspecification of the model (e.g., heteroscedasticity):

\begin{small}
<<eval = FALSE, echo = TRUE>>=
resids <- residuals(exampleglmmtree)
preds <- predict(exampleglmmtree)
plot(a_m_data$cluster, resids)
scatter.smooth(preds, resids)
@
\end{small}

The first plot did not indicate substantial variation in error variances across levels of the random effects. The second plot of fitted values against residuals also did not reveal a pattern indicating model misspecification.

@@

To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions. In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions. In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions. As the outcome variable was continuous in all simulated dataset, the GLMM tree algorithm and the trees resulting from its application will be referred to as LMM tree(s).

@@

\subsection{Study~I: Performance of LMM tree, LM tree and MERT in datasets with treatment-subgroup interactions}

@@

For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure~\ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct. Therefore, performance of MERT was only assessed in terms of predictive accuracy.

@@

Further splits were more accurately recovered by LMM trees yielding 90.40\% accuracy for the full partition comparted to only 61.44\% for LM trees. The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}. In the absence of random effects, LM and LMM tree were about equally accurate. In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate. When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$).

@@

For datasets with purely piecewise interactions, the same partition as in Study~I (Figure~\ref{fig:modelC}) was used. In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.

>>>>

\documentclass[doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{bigfoot}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}
\usepackage{Sweave}
\usepackage[caption=false]{subfig}

@@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^1$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees. Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods. Tree-based methods, also known as recursive partitioning methods, split observations repeatedly into groups so that they become increasingly similar with respect to the outcome within each group. Several tree-based methods take the mean of a continuous dependent variable or the majority class of a categorical dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}. Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.

Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3. As the observations in node~2 are not further split, node~2 is a terminal node. Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.

If the partition in Figure~\ref{fig:exampletree} would be used for prediction of a new observation, the new observation would be assigned to one of the terminal nodes according to its values on the splitting variables. The prediction is then based on the estimated distribution of the outcome variable within that terminal node. For example, the prediction may be the node-specific mean of a single continuous variable. In the current paper, we focus on trees where the terminal nodes consist of a linear (LM) or generalized linear model (GLM), in which case the predicted value for a new observation is determined by the node-specific parameter estimates of the (G)LM, while also adjusting for random effects.

@@

Tree-based methods are particularly useful for exploratory purposes, because they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, they are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov16, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15, AtheyImbe16}). Also, \citeA{ZhanyTsia12Bio} and \citeA{ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

@@

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyDoov16}, GLMM trees allow for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects \cite<e.g., >{HajjyBell11,SelaySimo12}, GLMM trees allow for treatment effect estimation, with continuous as well as non-continuous response variables.

The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions. In the second section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions. In the third section, we apply the GLMM tree algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. In the fourth and last section we summarize the results and discuss limitations and directions for future research. A tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material. In the tutorial, the artificial motivating dataset is used, allowing users to recreate the trees and models to be fitted in the next section.

@@

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. This dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Summary statistics of these variables are provided in Table~\ref{tab:sum_stats}. Each patient was part of one of ten clusters, each having a different value for the random intercept, which was generated from a standard normal distribution and uncorrelated with the partitioning variables.

\begin{table}[h!]
\small
\caption{\small Summary statistics for partitioning and outcome variables in the artificial motivating dataset}
\begin{tabular}{lcccc}
\thickline
 & min & max & mean & sd\\
\hline
depression & 3 & 16 & 9.12 & 2.66 \\
age & 18 & 69 & 45 & 9.56 \\
anxeity & 3 & 18 &  10.26 & 3.05 \\
duration & 1 & 17 & 6.97 & 2.90 \\
\hline
\end{tabular}
\label{tab:sum_stats}
\end{table}

The outcome variable was generated such that there are three subgroups with differential treatment effectiveness, corresponding to the terminal nodes in Figure~\ref{fig:exampletree}: For the first subgroup of patients (node~3) with short duration ($\le 8$) months of depressive symptoms prior to treatment and low anxiety scores ($\le 10$), Treatment~1 leads to lower post-treatment depression than in Treatment~2 (true mean difference = 2). For the second subgroup of patients (node~4) with short duration but high anxiety scores ($> 10$) post-treatment depression is about equal in both treatment conditions (true mean difference = 0). For the third subgroup of patients (node~5) with long duration ($> 8$ months) Treatment~2 leads to lower post-treatment depression than Treatment~1 (true mean difference = 2.5). Thus, duration and anxiety are true partitioning or moderator variables, whereas age is not. Anticipating the final results of our analyses, the treatment-subgroup interactions are depicted in Figure~\ref{fig:example_glimmertree}, which shows the GLMM tree that accurately recovered the treatment-subgroup interactions.

@@

The rationale behind MOB is that a single global GLM (or other parametric model) may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find better-fitting models in each cell of the partition. For example, to assess the effect of treatment, we may first fit a global GLM where the treatment indicator has the same effect/coefficient on the outcome for all observations. Subsequently, the data may be partitioned recursively with respect to other covariates, leading to separate models with different treatment effects/coefficients in each subsample.

@@

The MOB algorithm can be used to partition the dataset using additional covariates and find better-fitting local models. To this end, the MOB algorithm tests for parameter stability with respect to each of a set of auxiliary covariates, also called \textit{partitioning variables}, which we will denote by $U$. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find these partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) statistically test for parameter instability with respect to each of a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

@@

If at least one of the partitioning variables yields a $p$-value below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{k^*}$, the partitioning variable with the minimal $p$-value in step (2). The split point for $U_{k^*}$ is selected by taking the value that minimizes the instability as measured by the sum of the values of two loss functions, one for each of the resulting subgroups. In other words, the loss function is minimized separately in the two subgroups resulting from every possible split point and the split point yielding the minimum sum of the loss functions is selected. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a mixed-effects model would be more appropriate. The GLM in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a random-intercept only model, $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element corresponding to the random intercept for cluster $m$. For simplicity, we employ a cluster-specific intercept only, but further random effects can easily be included in $z_i$. Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$ and that the errors $\epsilon$ have constant variance across clusters. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

Although the random-effects part of the GLMM in Equation~\ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation~\ref{eq:fixedeffects_MOB} while still adjusting for random effects: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on subgroup $j$, but the random effects $b$ are global. To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects parameters are estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random-effects parameters.

@@

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the GLM tree is re-estimated in step (1) and the fixed- and random-effects parameters are re-estimated in step (2). Note that the random effects are not partitioned, but estimated globally. Only the fixed effects are estimated locally, within the cells of the partition. Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}. Typically, this converges if the tree does not change from one iteration to the next.

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. In addition to the treatment indicator, treatment outcome and the potential partitioning variables, the GLMM tree algorithm has also taken a random intercept with respect to the cluster indicator into account. As a result, the dependence between observations is taken into account, the true treatment subgroups have been recovered and the spurious split involving the anxiety variable no longer appears in the tree.

@@

@@

To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions. In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions. In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions. As the outcome variable was continuous in all simulated datasets, the GLMM tree algorithm and trees resulting from its application will be referred to as LMM tree(s).

@@

\subsection{Study~I: Performance of LMM trees, LM trees and MERTs in datasets with treatment-subgroup interactions}

@@

For MERT, the number of nodes and tree accuracy was not assessed, as the treatment-subgroup interaction design in Figure~\ref{fig:modelC} corresponds to a large number of regression tree structures, that would all be different but also correct. Therefore, performance of MERTs was only assessed in terms of predictive accuracy.

@@

Further splits were more accurately recovered by LMM trees yielding 90.40\% accuracy for the full partition comparted to only 61.44\% for LM trees. The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}. In the absence of random effects, LM and LMM trees were about equally accurate. In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate. When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$).

@@

To generate datasets with purely piecewise interactions, the same partition as in Study~I (Figure~\ref{fig:modelC}) was used. In other words, the outcome variable in this design was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.
